---
title: "Dialogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dialogue}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document, the dialogue with the AI
is described, including all control questions.
Of each question, the expected answer is
calculated.

```{r}
library(testthat)
```

The dialogue follows the skill levels of `[Spector & Ma, 2019]`
in increasing order:

- Interpretation
- Explanation
- Reasoning
- Analysis
- Evaluation
- Synthesis
- Reflection
- Judgement

## Expectations

A paper does not typically hold code (like an R markdown
document). Hence, in the end,
all the constants are copy-pasted into a paper
and then left unchecked.

To make it easy to compare the final paper with
the constants we've copy-pasted, here
are all the constants:


```{r expectations}
total_n_fish <- 300
n_fish_control <- 60
n_fish_transplanted <- 240
origin_values <- c("Lake", "Stream")
transplanted_values <- c("Lake", "Stream", "Control")
lowest_pre_mass <- 0.5
highest_pre_mass <- 3.6
n_survived <- 151
n_died <- 89
n_enclosures <- 80
n_enclosures_with_2_fish <- 3
n_enclosures_with_3_fish <- 74
n_enclosures_with_4_fish <- 3
anomous_enclosures <- c("L20", "L25", "L6", "L7", "S22", "S24")
cage_mass_mean_for_l1 <- 1.256667
cage_mass_stdev_for_l1 <- 0.4532475
cage_mass_mean_deviation_sd_for_l1 <- c(1.1546303, -0.5662836, -0.5883466)

abs_cage_mass_mean_deviation_sd_for_l1 <- abs(cage_mass_mean_deviation_sd_for_l1)

# Check consistency between these variables
expect_equal(
  total_n_fish,
  n_fish_control + n_fish_transplanted
)
expect_equal(
  n_fish_transplanted,
  n_survived + n_died
)
expect_equal(
  n_enclosures,
  n_enclosures_with_2_fish + n_enclosures_with_3_fish + n_enclosures_with_4_fish
)
expect_equal(
  length(anomous_enclosures),
  n_enclosures_with_2_fish + n_enclosures_with_4_fish
)
```

All solutions to the dialog are written here,
and tested below, during the dialogue.

## D1. Ask an AI to read the dataset

### D1.1: describe research paper conclusiom

Ask the AI:

```
I am going to upload data.

The data consist of fish in cages and their survival.

A paper that analysed this data concluded:
'survival is higher for atypically sized individuals within cages'.

How would you suggest to analyse the data?
```


### D1.2: read data

- The dialogue starts by uploading the data for the paper
  at [here](https://github.com/richelbilderbeek/Bolnick_and_Stutz_2017/blob/master/Bolnick_traits.txt).


```{r d1_read_data}
t_all <- read.csv("Bolnick_traits.txt", sep = " ")
```

### D1.3: first impression

```
I will now walk you through how the data should be understood.

A paper that analysed this data concluded:
'survival is higher for atypically sized individuals within cages'.

Do you already think that this data suggests that?
```

### D1.4: show first rows

Ask the AI:

```
Could you show the first rows of the data?
```

Expected:

```{r d1_show_first_rows}
knitr::kable(head(t_all))
```

We select only the relevant data here, so that the analysis can be checked
by humans more easily:

```{r d1_make_data_human_readable}
t <- t_all |> dplyr::select(fishID, origin, enclosure, pre_mass, transplant,  survived) |> dplyr::arrange(enclosure)
knitr::kable(head(t))
```

## D2. Describe the dataset

### D2.1: count fish

Ask the AI:

```
The `fishID` column denotes the ID of a fish. How many fish are in this dataset?
```

Expected:

```{r d2_count_fish}
expect_equal(total_n_fish, length(unique(t$fishID)))
```

The expected answer is `r total_n_fish`.

### D2.2: get the 'origin' values

Ask the AI:

```
The `origin` column denotes the location where each fish comes from.
What are the locations the fish come from?
```

Expected:

```{r d2_get_origin_values}
expect_equal(unique(t$origin), origin_values)
```

The correct answers is `r origin_values`.

### D2.3: get the 'transplant' values

Ask the AI:

```
The `transplant` column denotes the location where each fish is
transplanted to.
What are the locations the fish are transplanted to?
```

Expected:

```{r d2_get_transplant_values}
expect_equal(unique(t$transplant), transplanted_values)
```

The correct answers is `r transplanted_values`.

### D2.4: remove the controls and count the fish

Ask the AI:

```
Remove the rows for which the `transplant` value is 'Control'.
How many fish are left?
```

Expected:

```{r d2_remove_controls_and_count}
t_1 <- t |> dplyr::filter(transplant != "Control") 
expect_equal(n_fish_transplanted, length(unique(t_1$fishID)))
```

The correct answer is `r n_fish_transplanted`.

### D2.5: get the range of 'pre_mass'

Ask the AI:

```
The `pre_mass` column denotes the mass of a fish
before the transplantation.
What is the range of the values in this column?
```

Expected:

```{r d2_get_pre_mass_range}
expect_equal(lowest_pre_mass, min(t_1$pre_mass))
expect_equal(highest_pre_mass, max(t_1$pre_mass))
```

The correct answer is `r lowest_pre_mass` to `r highest_pre_mass`.

### D2.6: count the number of dead fish and fish that survived

Ask the AI:

```
The `survived` column denotes if the fish survived the experiment,
where 0 means it died and 1 means that it survived.
How many fish died? And how many survived?
```

Expected:

```{r d2_count_n_dead_and_n_survived}
expect_equal(n_survived, sum(t_1$survived == TRUE))
expect_equal(n_died, sum(t_1$survived == FALSE))
```

The correct answers are `r n_died` died and `r n_survived` survived.

### D2.7: count the number of enclosures

Ask the AI:

```
The `enclosure` column denotes the ID of an enclosure.
How many enclosures are in this dataset?
```

Expected:

```{r d2_count_enclosures}
expect_equal(n_enclosures, length(unique(t_1$enclosure)))
```

The correct answer is `r n_enclosures` enclosures.

### D2.8: count the number of fish in each enclosure

Ask the AI:

```
How many fish are in each enclosure?
```

Expected:

```{r d2_count_n_fish_per_enclosure}
n_fish_per_enclosure <- dplyr::count(t_1, enclosure)
expect_equal(n_enclosures_with_2_fish, sum(n_fish_per_enclosure$n == 2))
expect_equal(n_enclosures_with_3_fish, sum(n_fish_per_enclosure$n == 3))
expect_equal(n_enclosures_with_4_fish, sum(n_fish_per_enclosure$n == 4))
knitr::kable(head(n_fish_per_enclosure))
```

The correct answer is:

- `r n_enclosures_with_2_fish` enclosures have 2 fish
- `r n_enclosures_with_3_fish` enclosures have 3 fish 
- `r n_enclosures_with_4_fish` enclosures have 4 fish 

As a note to self, these are the anomalies:

```{r}
expect_equal(anomous_enclosures, n_fish_per_enclosure[n_fish_per_enclosure$n != 3, ]$enclosure)
knitr::kable(n_fish_per_enclosure[n_fish_per_enclosure$n != 3, ])
```

## **Q0**. Describe the conclusion

- A conclusion drawn from the data
  is that the extreme body masses
  are likelier to survive. Would you agree that the data supports
  this claim?

## D4. Reproduce the results in that paper

### D4.1: calculate 'cage_mass_mean'

Ask the AI:

```
To make a better comparison, we are going to standarize
body masses per enclosure.
Add a column to the
data called `cage_mass_mean` which holds the average
`pre_mass` within the enclosure each fish is in.
Could you show me the data for enclosure L1?
```

Expected:

```{r}
cage_mass_mean_per_enclosure <- t_1 |> 
  dplyr::select(enclosure, pre_mass) |>
  dplyr::group_by(enclosure) |>
  dplyr::summarise(cage_mass_mean = mean(pre_mass))
t_2 <- merge(t_1, cage_mass_mean_per_enclosure)
knitr::kable(head(t_2))
knitr::kable(t_2[t_2$enclosure == "L1", ])
expect_equal(
  t_2[t_2$enclosure == "L1", ]$cage_mass_mean,
  rep(cage_mass_mean_for_l1, 3),
  tolerance = 1.0e-6
)
```

For enclose L1, the expected  `cage_mass_mean` 
for each of the fish is `r cage_mass_mean_for_l1`.

### D4.2: calculate 'cage_mass_stdev'

Ask the AI:

```
To make a better comparison, we are going to standarize
body masses per enclosure. Add a column to the
data called `cage_mass_stdev` which is the standard
deviation of the `pre_mass` distribution of each enclosure
each fish is in.
Could you show me the data for enclosure L1?
```

Expected:

```{r}
cage_mass_stdev_per_enclosure <- t_2 |> 
  dplyr::select(enclosure, pre_mass) |>
  dplyr::group_by(enclosure) |>
  dplyr::summarise(cage_mass_stdev = sd(pre_mass))
t_3 <- merge(t_2, cage_mass_stdev_per_enclosure)
knitr::kable(head(t_3))
knitr::kable(t_3[t_3$enclosure == "L1", ])
expect_equal(
  t_3[t_3$enclosure == "L1", ]$cage_mass_stdev,
  rep(cage_mass_stdev_for_l1, 3),
  tolerance = 1.0e-6
)
```

### D4.3: calculate 'cage_mass_mean_deviation_sd'

Ask the AI:

```
To make a better comparison, we are going to standarize
body masses per enclosure. Add a column to the
data called `cage_mass_mean_deviation_sd`.
Its values are calculated per fish.
Each fish its `cage_mass_mean_deviation_sd`
equals the absolute difference between its `pre_mass`
and its enclosure's `cage_mass_mean`, 
divided by the `cage_mass_stdev` of its enclosure.
Could you show me the data for enclosure L1?
```

Expected:

```{r}
cage_mass_mean_deviation_sd_per_fish <- t_3 |> 
  dplyr::select(fishID, pre_mass, cage_mass_mean, cage_mass_stdev) |>
  dplyr::group_by(fishID) |>
  dplyr::mutate(
    cage_mass_mean_deviation_sd = 
      abs(pre_mass - cage_mass_mean) / cage_mass_stdev
  ) |>
  dplyr::select(fishID, cage_mass_mean_deviation_sd)
expect_equal(
  nrow(cage_mass_mean_deviation_sd_per_fish),
  n_fish_transplanted
)

expect_equal(
  nrow(t_3),
  n_fish_transplanted
)
t_4 <- merge(t_3, cage_mass_mean_deviation_sd_per_fish)
expect_equal(
  nrow(t_4),
  n_fish_transplanted
)

knitr::kable(head(t_4))
knitr::kable(t_4[t_4$enclosure == "L1", ])
expect_equal(
  t_4[t_4$enclosure == "L1", ]$cage_mass_mean_deviation_sd,
  abs_cage_mass_mean_deviation_sd_for_l1,
  tolerance = 1.0e-6
)
```

The values of `cage_mass_mean_deviation_sd` for enclosure `L1`
are expected to be `r abs_cage_mass_mean_deviation_sd_for_l1`.

### D4.4: reproduce the plot

Ask the AI:

```
Create a scatter plot of this data:
- Each point is a fish. 
- Use a blue color for fish
  that have 'Lake' as their origin.
- Use a green color for fish
  that have 'Stream' as their origin.
- On the X axis, put the values of `cage_mass_mean_deviation_sd`
  of each fish
- On the Y axis, put the `survived` of each fish.
- Put the fish in four categories:
  - Fish that have an `origin` of 'Lake' and
    a `transplant` of 'Lake' are in category 'LL'
  - Fish that have an `origin` of 'Lake' and
    a `transplant` of 'Stream' are in category 'LS'
  - Fish that have an `origin` of 'Stream' and
    a `transplant` of 'Lake' are in category 'SL'
  - Fish that have an `origin` of 'Stream' and
    a `transplant` of 'Stream' are in category 'SS'
- Show a trendline for each binomal fit on each of these 4 categories.
  - Use blue lines for categories that originate from a lake
  - Use green lines for categories that originate from a stream
  - Use solid lines for categories 'LL' and 'SS'
  - Use dashed lines for categories 'LS' and 'SL'
```

Expected is the plot in the paper:

```{r}
t_4$origin_transplant <- paste0(
  stringr::str_sub(t_4$origin,1 ,1),
  stringr::str_sub(t_4$transplant,1 ,1)
)
ggplot2::ggplot(
  data = t_4,
  ggplot2::aes(
    x = cage_mass_mean_deviation_sd,
    y = survived,
    color = origin_transplant,
    lty = origin_transplant
  )
) + ggplot2::scale_color_manual(
    values = c("blue", "blue", "green", "green")
  ) + 
  ggplot2::scale_linetype_manual(
    values = c("solid", "dashed", "dashed", "solid")
  ) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = FALSE
  )
```


## **Q1**: Ask if the conclusion is correct

- A conclusion drawn from the data
  is that the extreme body masses
  are likelier to survive. Would you agree that the data supports
  this claim?


## D5 Reproduce the results that show a flaw in the reasoning

### D5.1 Add relative standarized body mass

Ask the AI:

```
To make a better comparison, we are going to standarize
body masses per enclosure. Add a column to the
data called `cage_mass_mean_deviation_sd_rel`.
Its values are calculated per fish.
Each fish its `cage_mass_mean_deviation_sd_rel`
equals the difference between its `pre_mass`
and its enclosure's `cage_mass_mean`, 
divided by the `cage_mass_stdev` of its enclosure.
Could you show me the data for enclosure L1?
```

Expected:

```{r}
cage_mass_mean_deviation_sd_per_fish_rel <- t_4 |> 
  dplyr::select(fishID, pre_mass, cage_mass_mean, cage_mass_stdev) |>
  dplyr::group_by(fishID) |>
  dplyr::mutate(
    cage_mass_mean_deviation_sd_rel = 
      (pre_mass - cage_mass_mean) / cage_mass_stdev
  ) |>
  dplyr::select(fishID, cage_mass_mean_deviation_sd_rel)
expect_equal(
  nrow(cage_mass_mean_deviation_sd_per_fish_rel),
  n_fish_transplanted
)

expect_equal(
  nrow(t_4),
  n_fish_transplanted
)
t_5 <- merge(t_4, cage_mass_mean_deviation_sd_per_fish_rel)
expect_equal(
  nrow(t_5),
  n_fish_transplanted
)

knitr::kable(head(t_5))
knitr::kable(t_5[t_5$enclosure == "L1", ])
expect_equal(
  t_5[t_5$enclosure == "L1", ]$cage_mass_mean_deviation_sd_rel,
  cage_mass_mean_deviation_sd_for_l1,
  tolerance = 1.0e-6
)
```

### D5.2 Plot relative standarized body mass with same fit

Ask the AI:

```
Create a scatter plot of this data:
- Each point is a fish. 
- Use a blue color for fish
  that have 'Lake' as their origin.
- Use a green color for fish
  that have 'Stream' as their origin.
- On the X axis, put the values of `cage_mass_mean_deviation_sd_rel`
  of each fish
- On the Y axis, put the `survived` of each fish.
- Put the fish in four categories:
  - Fish that have an `origin` of 'Lake' and
     a `transplant` of 'Lake' are in category 'LL'
  - Fish that have an `origin` of 'Lake' and
     a `transplant` of 'Stream' are in category 'LS'
  - Fish that have an `origin` of 'Stream' and
     a `transplant` of 'Lake' are in category 'SL'
  - Fish that have an `origin` of 'Stream' and
     a `transplant` of 'Stream' are in category 'SS'
- Show a trendline for each parabolic fit on each of these 4 categories.
   - Use blue lines for categories that originate from a lake
   - Use green lines for categories that originate from a stream
   - Use solid lines for categories 'LL' and 'SS'
   - Use dashed lines for categories 'LS' and 'SL'
```

This should produce this plot:

```{r}
ggplot2::ggplot(
  data = t_5,
  ggplot2::aes(
    x = cage_mass_mean_deviation_sd_rel,
    y = survived,
    color = origin_transplant,
    lty = origin_transplant
  )
) + ggplot2::scale_color_manual(
    values = c("blue", "blue", "green", "green")
  ) + 
  ggplot2::scale_linetype_manual(
    values = c("solid", "dashed", "dashed", "solid")
  ) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    method = "lm", 
    formula = y ~ x + I(x^2),
    se = FALSE
  )
```

The differences are:
- this plot uses relative body masses, instead of absolute
- the fit is parabolic, instead of binomial, as
  the latter is 
  [assumed to be monotonically increasing/decreasing](https://github.com/richelbilderbeek/paper_critical_ai/issues/15).
  This means that a binomial distribution cannot be used to fit on data
  that is shaped like a `U` (as is assumed in the original paper: the extreme
  body masses have the highest fitness)

## **Q2**: Ask if the conclusion is correct

- A conclusion drawn from the data
  is that the extreme body masses
  are likelier to survive. Would you agree that the data supports
  this claim?
